---
title: "State-Level Happiness Analysis with Spatial Predictors"
editor: visual
format:
  html:
    embed-resources: true
    code-fold: true
execute:
  warning: false
  message: false
---

For this section, I wanted to see some other factors that might influence the happiness scores, including fiscal, educational, and geographic factors. As school resources, teacher and students' relationship, and education itself are allocation important variable in developing students' self esteem, I thought using education as a factor would be an interesting analysis. I used the fiscal data demonstrating states expenditures and the NAEP scores for 4th graders in the year 2022. By combining happiness scores with external datasets, the study investigates patterns in well-being and relationships with neighboring states.


```{r}
# Load libraries
library(tidyverse)
library(sf)
library(spdep)
library(viridis)
library(readxl)
library(rnaturalearth)
library(rnaturalearthdata)
library(corrplot)
library(readxl)
```

```{r}
# Load spatial and happiness datasets
us_states <- ne_states(country = "United States of America", returnclass = "sf")
happiness_state <- read_csv("../data/happiest_state_2022.csv", show_col_types = FALSE)

# Merge spatial and happiness data
merged_sf <- left_join(us_states, happiness_state, by = c("name" = "State"))
merged_sf <- merged_sf[!is.na(merged_sf$`Total Score`), ] # Remove rows with missing happiness scores


fiscal <- read_xlsx("../data/fiscal_predictors.xlsx")
outcomes <- read_xlsx("../data/outcome_data.xlsx")
dat2022 <- outcomes %>% filter(year == 2022)
performance_dat <- left_join(dat2022, fiscal, by = c('jurisdiction' = 'STABR', 'year' = 'TargetYear'))

merged_sf <- left_join(merged_sf, performance_dat, by = c("name" = "jurisLabel"))
merged_sf <- st_transform(merged_sf, crs = 3857)
```


### Visualizations

```{r}
library(tmap)

tm_shape(merged_sf) +
  tm_polygons("Total Score", palette = "Blues", title = "Happiness Score") +
  tm_layout(title = "State Happiness Scores")

```

This thematic map illustrates variations in happiness across states. Higher scores are concentrated in states like Hawaii, Maryland, and New Jersey, while lower scores are found in Louisiana and Arkansas.


### Spatial Autocorrelation

```{r}
library(spdep)

# Spatial neighbors and weights
neighbors <- poly2nb(merged_sf, queen = TRUE)
weights <- nb2listw(neighbors, style = "W", zero.policy = TRUE)

# Moran's I
morans <- moran.test(merged_sf$`Total Score`, weights, zero.policy = TRUE)
print(morans)

```

Moran's I was used to test for spatial autocorrelation in happiness scores. The global Moran's I is 0.280 with a  p-value of 0.001. This indicates significant clustering of happiness scores, with geographically close states tending to share similar levels of happiness.

### Predictors and Bivariate Relationships

```{r}
# Spatial lag of happiness scores
lagged_happiness <- lag.listw(weights, merged_sf$`Total Score`)

# Moran's I for math scores against lagged happiness scores
morans_bivariate <- lm.morantest(lm(math ~ lagged_happiness, data = merged_sf), weights)
print(morans_bivariate)

```

```{r}
library(ggplot2)

# Scatterplot
ggplot(data = merged_sf, aes(x = math, y = lagged_happiness)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(x = "Math Scores", y = "Lagged Happiness Scores", title = "Moran Scatterplot: Math vs Happiness")

```

```{r}
# Local Moran's I for math scores
local_morans_math <- localmoran(merged_sf$math, weights)

# Add LISA results for math to the spatial dataset
merged_sf$lisa_math <- local_morans_math[, "Ii"]

# Map clusters for math scores
tm_shape(merged_sf) +
  tm_polygons("lisa_math", palette = "RdBu", title = "LISA Clusters for Math") +
  tm_layout(title = "Local Spatial Autocorrelation (Math Scores)")

```

```{r}
# Spatial lag of happiness scores
lagged_happiness <- lag.listw(weights, merged_sf$`Total Score`)

# Moran's I for reading scores against lagged happiness scores
morans_bivariate <- lm.morantest(lm(reading ~ lagged_happiness, data = merged_sf), weights)
print(morans_bivariate)

```

```{r}
library(ggplot2)

# Scatterplot
ggplot(data = merged_sf, aes(x = reading, y = lagged_happiness)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(x = "Reading Scores", y = "Lagged Happiness Scores", title = "Moran Scatterplot: Reading vs Happiness")

```

```{r}
# Local Moran's I for reading scores
local_morans_reading <- localmoran(merged_sf$reading, weights)

# Add LISA results for reading to the spatial dataset
merged_sf$lisa_reading <- local_morans_reading[, "Ii"]

# Map clusters for reading scores
tm_shape(merged_sf) +
  tm_polygons("lisa_reading", palette = "RdBu", title = "LISA Clusters for Reading") +
  tm_layout(title = "Local Spatial Autocorrelation (Reading Scores)")

```

```{r}
# Spatial lag of happiness scores
lagged_happiness <- lag.listw(weights, merged_sf$`Total Score`)

# Moran's I for expenditure on education against lagged happiness scores
morans_bivariate <- lm.morantest(lm(TE11 ~ lagged_happiness, data = merged_sf), weights)
print(morans_bivariate)

```

```{r}
library(ggplot2)

# Scatterplot
ggplot(data = merged_sf, aes(x = TE11, y = lagged_happiness)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(x = "Total Expenditure on Education", y = "Lagged Happiness Scores", title = "Moran Scatterplot: Total Expenditure on Education vs Happiness")

```

```{r}
# Local Moran's I for expenditure on education
local_morans_total_exp <- localmoran(merged_sf$TE11, weights)

# Add LISA results for math to the spatial dataset
merged_sf$lisa_total_exp <- local_morans_total_exp[, "Ii"]

# Map clusters for total expenditure on education
tm_shape(merged_sf) +
  tm_polygons("lisa_total_exp", palette = "RdBu", title = "LISA Clusters for Total Expenditure on Education") +
  tm_layout(title = "Local Spatial Autocorrelation (Total Expenditure on Education)")

```

```{r}
# Spatial lag of happiness scores
lagged_happiness <- lag.listw(weights, merged_sf$`Total Score`)

# Moran's I for Local Revenues Student Activities against lagged happiness scores
morans_bivariate <- lm.morantest(lm(R1K ~ lagged_happiness, data = merged_sf), weights)
print(morans_bivariate)

```

```{r}
library(ggplot2)

# Scatterplot
ggplot(data = merged_sf, aes(x = R1K, y = lagged_happiness)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(x = "Local Revenues Student Activities", y = "Lagged Happiness Scores", title = "Moran Scatterplot: Revenues Student Activities vs Happiness")

```

```{r}
# Local Moran's I for Local Revenues Student Activitiese scores
local_morans_reading <- localmoran(merged_sf$R1K, weights)

# Add LISA results for Local Revenues Student Activitiese to the spatial dataset
merged_sf$lisa_reading <- local_morans_reading[, "Ii"]

# Map clusters for Local Revenues Student Activitiese scores
tm_shape(merged_sf) +
  tm_polygons("lisa_reading", palette = "RdBu", title = "LISA Clusters for Local Revenues Student Activitiese") +
  tm_layout(title = "Local Spatial Autocorrelation (Local Revenues Student Activitieses)")

```

```{r}
# Spatial lag of happiness scores
lagged_happiness <- lag.listw(weights, merged_sf$`Total Score`)

# Moran's I for math scores against lagged happiness scores
morans_bivariate <- lm.morantest(lm(E12 ~ lagged_happiness, data = merged_sf), weights)
print(morans_bivariate)

```

```{r}
library(ggplot2)

# Scatterplot
ggplot(data = merged_sf, aes(x = E12, y = lagged_happiness)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(x = "Math Scores", y = "Lagged Happiness Scores", title = "Moran Scatterplot: Math vs Happiness")

```

```{r}
# Local Moran's I for happiness scores
local_morans <- localmoran(merged_sf$E12, weights)

# Add LISA results to the spatial dataset
merged_sf$lisa <- local_morans[, "Ii"]

# Map clusters
library(tmap)
tm_shape(merged_sf) +
  tm_polygons("lisa", palette = "RdBu", title = "LISA Clusters for Expenditures on Employee Benefits") +
  tm_layout(title = "Local Spatial Autocorrelation (Expenditures on Employee Benefits)")

```

```{r}
library(sf)
library(tmap)
library(dplyr)

# Ensure merged_sf is projected in a suitable CRS (e.g., EPSG:3857 for meter units)
merged_sf <- st_transform(merged_sf, crs = 3857)

# Create a 50 km buffer around state borders
state_buffers <- st_buffer(merged_sf, dist = 1000)

# Visualize the states and their buffers
tm_shape(merged_sf) +
  tm_polygons("Total Score", palette = "Blues", title = "Happiness Scores") +
  tm_shape(state_buffers) +
  tm_borders(col = "red") +
  tm_layout(title = "50 km Buffers Around States")

# Find intersections (neighbors within the buffer)
neighbors_in_buffer <- st_intersects(state_buffers, merged_sf)

# Add neighbor happiness scores to each state's data
merged_sf <- merged_sf %>%
  mutate(
    neighbor_happiness_mean = sapply(neighbors_in_buffer, function(neighbors) {
      if (length(neighbors) > 1) {
        mean(merged_sf$`Total Score`[neighbors], na.rm = TRUE)
      } else {
        NA
      }
    })
  )

# Compare happiness scores
comparison <- merged_sf %>%
  st_drop_geometry() %>%
  select(name, `Total Score`, neighbor_happiness_mean) %>%
  arrange(desc(`Total Score`))

# Display the comparison table
print(comparison)

# Visualize neighbors' average happiness scores
tm_shape(merged_sf) +
  tm_polygons("neighbor_happiness_mean", palette = "-RdYlBu", title = "Neighbor Happiness Mean") +
  tm_layout(title = "Average Happiness of Neighboring States Within 1 km")

```


Math and Reading Scores:

Moran's I tests and scatterplots were used to examine the relationship between educational performance (e.g., math and reading scores) and happiness.
Results for math and reading showed no significant spatial clustering in their relationship with happiness (p > 0.05).

Fiscal Predictors:

Relationships between happiness scores and expenditures (e.g., total educational spending and local revenues) were tested.
Local revenues for student activities showed significant spatial clustering (Moran's I = 0.279, p < 0.001).
Other fiscal predictors (e.g., expenditures on employee benefits) exhibited no significant spatial patterns.

Buffer Analysis:

States with higher happiness scores tended to have neighbors with above-average happiness. For example, Hawaii scored highest but lacked direct neighbors, while Maryland and New Jersey showed high scores alongside supportive neighbors.
